---
title: "miRNA Fetal Tissues: miRNA expression"
author: Nikita Telkar
output:
  html_document:
    keep_md: yes
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    theme: cosmo
    highlight: zenburn 
#tango, pygments, kate, monochrome, espresso, zenburn, haddock, textmate.
---

***
|object name | description |  
|--|--|
|`pDat`| Phenotype characteristics of the samples|
|`t_only_miRNA`| Expresssion only data --> Transposed version of `raw_miRNA-df`|  
|`meta_dat`|Manipulated data  |

***

## **Loading + Transforming Data**

Loading libraries:  
```{r, message=FALSE, echo =TRUE}
library(dplyr)
library(tidyr)
library(readxl)
library(tidyverse)
library(DT)
library(knitr)
library(kableExtra)
library(ggplot2)
library (plotly)
library(rmarkdown)
library(limma)
library (xlsx)
raw_miRNA <- read_excel("Z:/Nikita/Projects/mirna_fetal_tissues/data/external/magda_cleanedmiRNAdata.xlsx", col_names = FALSE, sheet = "miRNA")
``` 

```{r setup, include=FALSE}
opts_chunk$set(error = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=40),tidy=TRUE)

```
***
Transposing data:
```{r}
t_raw_mirna <- t(raw_miRNA)
colnames(t_raw_mirna) = t_raw_mirna[1, ]
t_raw_mirna <- t_raw_mirna[-1, ] #matrix
raw_miRNA_df <- as.data.frame(t_raw_mirna)
colnames(raw_miRNA_df)[1] <- "ID"
``` 

*** 
## **miRNA Names**  

Trying to get a list of only the miRNA names:  
```{r, error=TRUE}
mi_names <- raw_miRNA_df %>% 
  select(1)

select (raw_miRNA_df)

```  
The above error and `select` shows that there are duplicates of several miRNA names -- this will cause problems later on. So, it's better to have an object/value containing the names as a list which might be of use later.   
Using `t_raw_mirna` where the names are in separate columns as opposed to rows as above in `raw_miRNA_df`:  

```{r}
mi_names <- colnames(t_raw_mirna) 
mi_names <- tail (mi_names, -28)
head (mi_names)
```  
This has now given us a variable `mi_names` which is a list of the names of the miRNAs as they appear in the raw data `raw`.  

## **Supplementary Processed Data**  

```{r}
pDat <- raw_miRNA_df[ c(1:28) ]

pDat <- head(pDat,-8)

#o_miRNA.df <- raw_miRNA_df [-c(2:28)]

only_miRNA <- as.data.frame(raw_miRNA_df)
only_miRNA <- head (only_miRNA, -8)
only_miRNA <- only_miRNA [-c(1:28)]
t_only_miRNA <- t(only_miRNA)
t_only_miRNA <- as.data.frame(t_only_miRNA)

#Adding Column Names:
id <- pDat %>% 
  select ("ID")

id <- unlist(id) %>% 
       as.character(id)

colnames(t_only_miRNA) <- id
```

***

## **Analysis**  
Converting data from factors to numeric:

```{r, results=FALSE}

t_only_miRNA <- t_only_miRNA %>% 
                  mutate_all(as.character) %>% 
                  mutate_all(as.numeric)
t_only_miRNA
```
***

## **Calculating Mean Expression**  

Before adding in any character vectors, I'm going to do a numeric calculation to get the mean expression-   

To calculate the <span style='font-weight:bold;font-family:"Palatino Linotype";font-size:17px;l'><span style='color:#689127;'>mean expression</span></span> of the miRNAs across all 106 samples:  

```{r}
meta_dat <- t_only_miRNA %>% 
  mutate(Mean_expression = rowMeans(t_only_miRNA))

meta_dat <- meta_dat %>% 
  select (Mean_expression, everything())
```
***

**However**, applying the above step \ref{convert} `mutate` leads to `tidyverse` dropping the rownames. This is where the earlier `mi_names` list will come in hnady.  
And so, we can either have the row names as actual `rownames` or as a separate data column.  
--> Trying out as actual `rownames`:  

```{r, error=TRUE, echo =FALSE, results = 'hide', message=FALSE, warning= FALSE}

rownames(t_only_miRNA) <- mi_names

```  
  
  
The above doesn't work since there are duplicate miRNAs variables since expression was observed more than once from a different genomic location  
--> So,  adding `names` as a separate column  

```{r}

meta_dat <- meta_dat %>% 
                  mutate (names = mi_names) #adding miRNA names as a column


meta_dat <- meta_dat %>% 
                  select('names', everything())

#print(as_tibble(t_only_miRNA), n = 5)

paged_table(meta_dat)
```



```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

Code didn't work:

sub_ti <- sub_ti %>% 
  rownames_to_column() %>% #converts the rownames as a separate data column
  mutate_if(is.factor, function(x)as.numeric(as.character(x))) #specifying is.factor, since rownames are as.character

sub_ti <- sub_ti %>% 
  mutate (Mean_expression = rowMeans(sub_ti))

```

***
Chromosome information added in the same manner as above, `chr`:  

```{r}

chr <- raw_miRNA %>% 
        select(108)

chr <- tail(chr, -28)

chr <- unlist(chr)

meta_dat <- meta_dat %>% 
                  mutate (chr = chr) #adding chr as a column


meta_dat <- meta_dat %>% 
                  select('chr', 'names', everything())

```
***

Joining the `chr` column with the miRNA `names` column to give unique characters (since as seen before, there are duplicates):  
```{r}

meta_dat <- meta_dat %>% 
    mutate(row_names = paste(chr, names, sep="_"))

up_rn <- meta_dat %>% 
  select ("row_names")

up_rn <- unlist(up_rn) %>% 
          as.character(up_rn)

row.names(meta_dat) <- up_rn

```  
### **Further Duplication Problem**
This gives the same error as before --> there are still some duplicates present.  

```{r}

#duplicated(up_rn) #duplicates present in true or false
#unique(up_rn) #unique values present

meta_dat %>% 
    summarise(count = n_distinct(row_names))
```
The above gives the number of unique elements in the row_names column.   
Meaning that:  
Of the 2794 values, 2666 are unique.  
i.e. 2794 - 2666 = 128 are duplicates  

Removing the duplicated names then:  

```{r}

#dups <-meta_dat %>% 
#        group_by(row_names) %>% 
#       filter(n()>1)

meta_dat <- meta_dat %>% 
            distinct(row_names, .keep_all = TRUE) #removing duplicates based on row_names

up_rn <- meta_dat %>% 
  select ("row_names")

up_rn <- unlist(up_rn) %>% 
          as.character(up_rn)

row.names(meta_dat) <- up_rn

paged_table(meta_dat)
```


```{r, include=FALSE, eval = FALSE}

Code didn't work: 

meta_dat <- meta_dat %>% 
              mutate_if(is.numeric, (Mean_expression = rowMeans(meta_dat)))

meta_dat <- meta_dat %>% 
              mutate_if(is.numeric, function(x)Mean_expression = rowMeans(x))

meta_dat <- meta_dat %>% 
              mutate_if(is.numeric, funs(Mean_expression = rowMeans(meta_dat)))


meta_dat <- meta_dat %>% 
              select(everything()) %>% 
              mutate_if(is.numeric, Mean_expression = rowMeans(meta_dat))

```





```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

Code not run:

sub_ti <- t_only_miRNA [c(1:4)]
sub_ti <- head(sub_ti, 10)
col.names <-1:4
sub_ti <- setNames(sub_ti,col.names)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
Code didn't work:

sub_ti <- as.numeric(sub_ti)


apply(sub_ti, 1, mean)

sub_ti$mean <- rowMeans(sub_ti, na.rm=TRUE)

rowMeans(sub_ti)


sub_ti %>% 
  mutate(avg = mean(select_all(sub_ti)))

```

```{r, include = FALSE, eval=FALSE}
is.data.frame(sub_ti) #test for which type of data 

```



```{r, eval=FALSE, include=FALSE}
#check max expression --> select for miRNA's and make a plot

t_only_miRNA %>% 
            filter(Mean_expression == max(Mean_expression))

max_miRNA <- t_only_miRNA %>% 
          # count(Mean_expression > 15)
              filter(Mean_expression > 19.5) %>% 
              arrange((desc(Mean_expression)))

ggplot(data = t_only_miRNA, aes( x = names, y = Mean_expression)) +
  geom_bar(stat = "fill",  subset=.(Freq>4)))
          

exp <- t_only_miRNA %>% 
  select (Mean_expression) %>% 
  mutate(Exp_0 = frequency(Mean_expression ==0), Exp_more_15 = Mean_expression > 15)

ggplot(data = exp, aes(x = Exp_more_15,)) +   #get frequency of each for 0, 10-15, 15-20, 20+
      geom_bar (stat = "count") 

```
***

`Mean_expression` was rounded to give a whole number `r_exp` to make it easier for distinguising expression levels:  

```{r}
#frequency table

meta_dat <- meta_dat %>% 
        mutate (r_exp = round(Mean_expression))

meta_dat <- meta_dat %>% 
    select ('chr', 'names', 'Mean_expression', 'r_exp', everything())
```  
The above steps (i.e. any further data manipulation step) removes the row names. So, we'll have to leave adding the row names to the last step, just before saving the data. But till then, I'll make the row_names column the first one so that the data is easier to identify.  

```{r}

meta_dat <- meta_dat %>% 
    select ('row_names', everything())
```

***  
3 separate variables were created having the frequencies (no. of miRNAs) according to expression levels, and then put together in a separate df, `exp_data`   

Separating the novel miRNA according to mean expression levels:  
1. `zero` <- Average expression level = 0  
2. `bet_1_10` <- Average expression levels from 1 to 10  
3. `more_10` <- Average expression level > 10  

```{r,warning=FALSE}
meta_dat %>% 
  group_by(r_exp) %>% 
  summarise(freq=n())

zero <- meta_dat %>% 
  filter(r_exp == 0) %>% 
  summarise(zero_exp=n())

bet_1_10 <- meta_dat %>% 
  filter(between(r_exp, 1, 10)) %>% 
  summarise(mid_exp = n())
  
more_10 <- meta_dat %>% 
  filter(r_exp > 10) %>% 
  summarise(more_exp = n())

#zero + bet_1_10 + more_10

exp_data <- bind_cols(zero, bet_1_10, more_10)
rn <- colnames(exp_data)

exp_data <- t(exp_data)
exp_data <- as_tibble(exp_data) %>% 
  mutate(exp_levels = rn)

colnames(exp_data)[1] <- "number"

```

A **bar plot** showing the number of novel miRNA according to their mean expression levels was created using  
<span style='font-weight:bold;font-family:"Palatino Linotype";font-size:17px;'><span style='color:#689189;'>ggplot2</span></span> and <span style='font-weight:bold;font-family:"Palatino Linotype";font-size:17px;'><span style='color:#689128;'>plotly</span></span>:  

```{r}
a <-exp_data %>%
  mutate(exp_levels = fct_relevel(exp_levels, c('zero_exp', 'mid_exp', 'more_exp'))) %>%
  
  ggplot(aes(x = exp_levels, y = number, fill = exp_levels)) +
  geom_bar(stat = "identity", width=0.5, fill="#FFC7EA", colour = "black") +#text = paste0(exp_levels, "Avg expression")) + #default value
  
  geom_text(aes(label = number), nudge_y = 50) +
  scale_x_discrete(labels = c('Zero', 'Mid (1-5)', 'High')) +
  labs(title="Expression Levels of novel miRNAs", subtitle = "Average expression levels of 2794 miRNAs", x ="Expression Level", y ="No. of miRNA") +
  coord_cartesian(ylim=c(0,2000))

a
```
***

```{r, include=FALSE}
a
#text <- c('Avg expression level = 0', 'Avg expression levels from 1 to 10', ' Avg expression level > 10')

#ggplotly(a) %>% #(tooltip = c('Avg expression level = 0', 'Avg expression levels from 1 to 10', ' Avg expression level > 10')) 
 # layout(title = list(text = paste0('<br>',
  #                                  'Expression Levels of novel miRNAs',
   #                                 '<br>',
    #                                '<sup>',
     #                               'Average expression levels of 2794 miRNAs',
      #                              '</sup>')))

```
***

**Finally Adding Row Names**

```{r}
row.names(meta_dat) <- up_rn
```  
***

## **Saving Data**  

```{r, error=FALSE, eval=FALSE}

eDat <- meta_dat [c(6:111)] #expression only data

write.xlsx2(eDat, file ="Z:/Nikita/Projects/mirna_fetal_tissues/data/processed/miRNA_fetal_tissues_eDat.xlsx", col.names = TRUE, row.names = TRUE, append = FALSE)  

write.xlsx2(pDat, file ="Z:/Nikita/Projects/mirna_fetal_tissues/data/processed/miRNA_fetal_tissues_pDat.xlsx", col.names = TRUE, row.names = TRUE, append = FALSE)  

write.xlsx2(meta_dat, file ="Z:/Nikita/Projects/mirna_fetal_tissues/data/processed/miRNA_fetal_tissues_meta_dat.xlsx", col.names = TRUE, row.names = TRUE, append = FALSE)  

#write.xlsx for smaller files and write.xlsx2 for bigger files
```




