---
title: "miRNA Fetal Tissues: Data Exploration"
author: Nikita Telkar
output:
  html_document:
    keep_md: yes
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    theme: cosmo
    highlight: zenburn 
#tango, pygments, kate, monochrome, espresso, zenburn, haddock, textmate.
---



## <span style="color:maroon">Data Exploration</span>

**Read the data in using readxl::read_xlsx. What other ways can we read this data in?**  
Data can be read in depending upon the file type. So, for an excel file, we can use -  
read.excel  
read.xls  
read.table

***
**What type of data is in the magda_cleanedmiRNAdata.xlsx file? How is it structured?**  
The data shows micro RNA levels detected from different tissues obtained at different gestational ages of several
differnt foetuses.  



***

**How many “variables” are present?**  
There are 27 variables present, namely -  
Reading data in  
```{r, message=FALSE, echo =TRUE}
library(dplyr)
library(tidyr)
library(readxl)
library(tidyverse)
library(DT)
library(knitr)
library(kableExtra)
library(ggplot2)
library(rmarkdown)
raw_miRNA <- read_excel("Z:/Nikita/Projects/mirna_fetal_tissues/data/external/magda_cleanedmiRNAdata.xlsx", col_names = FALSE, sheet = "miRNA")
```  
Variables present:  

```{r}
print(raw_miRNA[2:28, 1], n = 27)
```


```{r, eval = FALSE, message=FALSE, echo=FALSE} 
#var <- raw_miRNA[2:28, 1]        #alt way to print --> convert into separate variable and print
#print(as_tibble(var), n = 27)
```

***


**How many patients are present?**  
There are a total of 106 patients present, of which  

**45 are female:**  
```{r,warning=FALSE, collapse=TRUE}
raw_miRNA %>% 
  str_extract_all('FEMALE') %>%
  unlist %>%
  length
```

**61 are male:**  
```{r, warning=FALSE, collapse=TRUE}
raw_miRNA %>% 
  str_extract_all('MALE') %>%
  unlist %>%
  length
```  
Hence,  
106 (word: MALE) - 45 (word: FEMALE)  
= **61**  

***

**How many samples are present?**  
- There are 106 total patients.  
- Number of miRNAs =  
2822 (total rows) - 29 (variables plus initial empty rows)  
= 2793  
Hence,  
2793 * 106  
= 296,058

***
**What is the breakdown of tissues / trimester? Generate a pretty table using kableExtra::kable(). Do this with at least 1 other variable of your choosing.**  
I decided to subset 5 variables -  
- tissue  
- extended sample condition information  
- gestational age  
- trimester  
- sex  

```{r}
sub_miRNa <- raw_miRNA[c(5,7,8,9,10),]
t_sub_miRNA <- t(sub_miRNa)
```   

```{r,echo=FALSE}
colnames(t_sub_miRNA) = t_sub_miRNA[1, ]
t_sub_miRNA <- t_sub_miRNA[-1,]
t_sub_miRNA <- na.omit(t_sub_miRNA)
```  
***
## **Summary Characteristics**  
**Summary of the variables: **

```{r}
summary(t_sub_miRNA)
```  
```{r, echo=FALSE}
datatable(t_sub_miRNA)
```


```{r, eval=FALSE, echo=FALSE}
unique(t_sub_miRNA[,1])  #To list out the the differnt tissue types
```    
***

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#tissue.trimester <- t_sub_miRNA %>% 
#+   group_by (tissue) %>% 
#+   summarise(trimester)
#Error in UseMethod("group_by_") : 
#  no applicable method for 'group_by_' applied to an object of class "c('matrix', 'character')"
  
  
  
#  summarise(.t_sub_miRNA)
#Error in summarise(.t_sub_miRNA) : object '.t_sub_miRNA' not found



#(t_raw3[,1]) %>% 
#+ filter(!is.na) %>% 
#+ summarise(Unique_Elements)
#Error in UseMethod("filter_") : 
#  no applicable method for 'filter_' applied to an object of class "character"
```
**Since the data was all in a matrix form before, I converted it into a dataframe `raw_miRNA.df`: **

```{r}
t_raw_mirna <- t(raw_miRNA)
colnames(t_raw_mirna) = t_raw_mirna[1, ]
t_raw_mirna <- t_raw_mirna[-1, ]
raw_miRNA.df <- as.data.frame(t_raw_mirna)
```  

**`raw_miRNA.df` is now:  **
1. the transposed version of `raw_miRNA`  
2. a dataframe  
3. has column names  
4. easier to view and understand  

***

```{r, echo = FALSE, message=FALSE, warning=FALSE, eval=FALSE}
tissue.trimester <- raw_miRNA.df %>% 
+   group_by (tissue) %>% 
+   summarise(trimester)
```

```{r, echo = FALSE, message=FALSE, warning=FALSE, eval=FALSE}
for next time: separate initial raw_miRNA.df by variable, and miRNA data. Since some miRNA are found in more than 1 chr and are duplicated in column name --> after transposing from raw to df

figure out duplicates of miRNA, maybe categorize by chr?
```

**Moving all of the non-miRNA data (characteristics) into a separate df:  **
```{r}
pDat <- raw_miRNA.df[ c(1:28) ]
colnames(pDat)[1] <- "ID"

pDat <- head(pDat,-8)

pDat <- as_tibble(pDat)
paged_table(pDat)
```
 
**And only the miRNA expression level data into `o_miRNA.df` **

```{r error=FALSE, message=FALSE, warning=FALSE, tidy=TRUE}

o_miRNA.df <- raw_miRNA.df [-c(2:28)]

paged_table(o_miRNA.df)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
#trying to change headers - Required for next analysis step: header rames (miRNA names) changed automatically as duplicates
o_miRNA.df <- raw_miRNA.df
names(o_miRNA.df)[1] <- "x"
o_miRNA.df <- as_tibble(o_miRNA.df)
o_miRNA.df <- o_miRNA.df [-c(2:28)] #same as above

o_miRNA.df <- raw_miRNA.df
colnames(o_miRNA.df) = o_miRNA.df[108, ] #weird names. 
o_miRNA.df <- o_miRNA.df [-c(2:28)]

```
***

## **Tissue versus Trimester**  

**For calculating the breakdown of tissues / trimester: **  


```{r, error=FALSE, message=FALSE}

tri_tis <- (pDat [(c(5,9))])

tri_tis <- rename(count(tri_tis, trimester, tissue), Freq = n) #getting frequency (sum) of all total number of samples

tri_tis <- tri_tis[, c(2,1,3)]

tri_tis <- tri_tis %>% 
            spread(trimester, Freq) %>% 
            replace(is.na(.), 0)

names(tri_tis) <- c("Tissue","Trimester_1", "Trimester_2", "Trimester_3")

#tri_tis %>%
#  mutate (Tissue = cell_spec(Tissue, color = "pink", background ="black")) %>% 
  kable(tri_tis, align = "c") %>% 
    kable_styling(bootstrap_options = c("striped", position = "centre", font_size = 12))
    
```

All the tissues except for **chorionic villi** contained samples from only trimester 2 pregnancies.  

## **Bar Plot**

**Showing the above as a bar plot:**

```{r}
  ggplot (data = pDat, aes(x = trimester, fill=tissue)) + #max no of samples without specified y-axis legth
   labs(title="Tissues per Trimester", subtitle = "Colors denote count of tissue", x ="Trimester", y ="No. of samples") +
  # scale_fill_brewer(palette=11) + #to add colours from a shaded pallete, only works for multiple variables
   geom_bar (stat = "count", width=0.8) +
   coord_cartesian(ylim=c(0,100))
    
````






