---
title: "Novel miRNAs"
author: "Nikita Telkar"
date: "December 16, 2019"
output: github_document
--- 
 
## Data Loading and Visualisation:  

```{r, message=FALSE, echo =TRUE} 
library(rmarkdown)
# library(installr)
library(dplyr) 
library(tidyr) 
library(readxl) 
library(DT) 
library(knitr) 
library(kableExtra) 
library(ggplot2)
library(tidyverse)
library(devtools)
library(magrittr)
library(arsenal)
library(xlsx)
library(tinytex)
```
***
Reading in the `novel` and the `predicted` miRNA files:  

```{r, error=FALSE, message=FALSE}
raw <- read_excel("Z:/Nikita/Projects/snc_placenta/data/raw/quantification_novel_mirnas_rpm_norm.xlsx", col_names =FALSE)

pred <- read_excel("Z:/Nikita/Projects/snc_placenta/data/raw/predictions.xlsx", col_names = TRUE)

pData <- read_excel("Z:/Nikita/Projects/mirna_fetal_tissues/data/processed/miRNA_fetal_tissues_pDat.xlsx", col_names = TRUE)
```

```{r, layout="l-body-outset"}
print(pred, n = 15)
```
***

### Selecting Rob Lab Samples:  

Only want to analyze Rob lab samples, so:  
```{r}
rob <- raw [1:32] #dropping all TCGA samples and keeping only Roblab samples
colnames(rob) = rob[1, ]
rob <- rob [-1,]
```

***
Joining `rob` and `pred`:  

```{r}

#fullJoinDf <- full_join(tableA,tableB,by=”Customer.ID”)
#band_members %>% full_join(band_instruments2, by = c("name" = "artist"))")

same <- pred %>% 
              inner_join(rob, by = c ("precursor_name" = "precursor")) #joining and keeping only common columns from pred and rob

#Doesn't work because rob has teo predicted miRNAs with the same precursor name

#same <- na.omit(same)

#same <- same %>%
 #        drop_na()
  
#same <- unique(same)

#same <- same %>% 
         # distinct(precursor_name, .keep_all = TRUE) #makes no sense as two of each with the same precursor name
```

***
```{r, eval = FALSE, include=FALSE}

All reads that were not aligned with a known miRNA sequences (mirBase vXX) were assessed an implemntation of the miRDeep2 (a commonly used tool in the characterization of novel miRNA sequences [refs]) deployed in the mirmaster platform. Prospective miRNA precursors were filtered according to their expression level (>= 1 RPM in at least 10% of samples), their novomirank score (<= 1.5) and their predicted probability of being a true precursor (>= 65%) [28911107].  
```

### Subsetting Data:  

According to the [paper](https://academic.oup.com/nar/article/44/6/e53/2499429) describing how scoring of novel miRNAs takes place, the `novomirank_score` variable calculated takes into consideration 24 different factors such as precursor length, percentage of the 4 differnet bases in the precursor, overlapping precursors. It returns a z-score (maximum threshold score of 3) after comparing with all the entries present in the latest version of the [novoMiRank database](https://ccb-compute2.cs.uni-saarland.de/novomirank/); higher the score, higher the chance of it being novel.  

And so, selecting for only `Novel` entries, and then segregating based on the score:
```{r}
pred_nov <- same %>% 
            filter(Category =="Novel") %>% 
        #   distinct(precursor_name, .keep_all = TRUE) %>% 
        #   arrange(desc('novomirank_score')) %>% 
        #   filter(novomirank_score > 1.5 )
            mutate_at(vars(novomirank_score), round, digits = 1)

nov_rank <-pred_nov %>%
  group_by(novomirank_score) %>% 
  summarise(number = sum(novomirank_score))

#maybe make a new object and then plot? 
nov_rank %>%   
  ggplot(aes(x=factor(novomirank_score), y=number, fill=number)) + #using x=factor()to display all x-axis values
  geom_bar(stat = "identity") +
  scale_fill_continuous(low="coral", high="purple") +
  labs(title ="Number of Novel miRNAs by NovoMiRank Score", subtitle ="The NovoMiRank score has threshold values of 0 to 3",x = "NovoMiRank Score (z-score)", y = "Number of miRNAS") 

```  

Now further plotting the `Novel` miRNAs by their predicted probablity of having a **true precursor**:  

```{r}


pred_score <-pred_nov %>%
  mutate_at(vars(probability), round, digits = 2) %>% 
  mutate(prob_percent = probability * 100)

pred_score <- pred_score %>%
  group_by(probability) %>% 
  summarise(number = sum(probability))

pred_score <-pred_score %>%
  mutate_at(vars(number), round)

pred_score <- pred_score %>% 
  mutate_at(vars(probability), funs(probability * 100)) %>% 
  mutate_at(vars(probability), funs(paste(probability, "%", sep ="")))

table(pred_score)
```  

Depicting the above table as a plot:  

```{r}
pred_score %>%   
  ggplot(aes(x=probability, y=number, fill=number)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_continuous(low="coral", high="purple") +
#  scale_x_continuous(breaks = c(50, 60, 70, 80, 90, 100))
#  scale_y_discrete(limits=c(50, 100))
  coord_flip() +
  labs(title ="Probability of Novel miRNAs having a true precursor", x = "Probability depicted as Percentages", y = "Number of miRNAS") 
```
***

### Matching Sample Names and miRNAs:  


```{r}
pDat <- pData %>% 
  mutate_at(vars(gscID), funs(str_replace_all(gscID, "_" , "-")))

t_rob <- t(rob)
t_rob <- as.data.frame(t_rob)
t_rob <- t_rob %>% 
  rownames_to_column("VALUE") #convert rownames to first column

match <- t_rob %>% 
  inner_join(pDat, by = c("VALUE" = "gscID"))

pDat <- match[c(1,28901:28927)] #updated pDat for the Rob lab samples only
colnames (pDat) [1] <- "sample"

match <- match[-c(28900:28927)]

colnames(match) = t_rob[2, ] #doesn't work for some ungodly reason. Unlisting it, as.character doesn't work either

mi <- t_rob[2, ] #So converting to separate object as rows

mi <- t(mi) #converting the row to a column

is.character(mi)

colnames(match) = mi[, 1] #adding the column as column names

eDat <- match #Roblab samples miRNA expression data

eDat <- eDat %>% 
  mutate_at(vars(-miRNA), funs(as.character)) %>% 
  mutate_at(vars(-miRNA), funs(as.numeric))

t_edat <- data.frame(t(eDat))
colnames(t_edat) <- t_edat[1,] #again the idiotic reason as above
cn <- t_edat[1,]
cn <-t(cn)
colnames(t_edat) <- cn[,1]
t_edat <- t_edat [-1, ]

colnames (eDat) [1] <- "sample"

rm(match)

#str(eDat)

```
***  

### Generating `meta_dat`:  

```{r}
meta_dat <- pDat %>% 
  inner_join(eDat, by = "sample")

meta_dat %>% 
  count(sex)

cn <- colnames(meta_dat)

t_meta <- rbind(cn, meta_dat)
t_meta <- t(t_meta)
colnames(t_meta) = t_meta[1, ]
     
meta_dat <- meta_dat %>%
  mutate(sex2 = recode(sex,"MALE" = 1, "FEMALE" = 2)) %>% #recoding sex to numeric values
  select(c(1:10), "sex2", everything())
  
meta_dat %>% 
  count(sex2)

```
***

### Saving Data:  

```{r, eval=FALSE}

options(java.parameters = "-Xmx1000m")

write.xlsx2(t_edat, file ="Z:/Nikita/Projects/snc_placenta/data/processed/snc_placenta_rob_eDat.xlsx", col.names = TRUE, row.names = TRUE, append = FALSE) #

write.xlsx2(pDat, file ="Z:/Nikita/Projects/snc_placenta/data/processed/snc_placenta_rob_pDat.xlsx", col.names = TRUE, row.names = TRUE, append = FALSE)  


```  


